Stochastic Newton with Arbitrary Sampling

### 2023-05-24 paper/Stochastic Newton with Arbitrary Sampling.tex
Added russian

	In practice, various sampling strategies are used for SGD-type algorithms to improve further the performance. One of the most famous sampling mechanisms is so-called Importance Sampling \cite{gower2019sgd, https://doi.org/10.48550/arxiv.1401.2753, 9413313}. The idea is to compute gradients of the functions that have more impact on the problem. Another method of random optimization is Random Reshuffling, which does iterative gradient steps passing reshuffled data \cite{mishchenko2020random}. \cite{richtarik2016parallel} studies many other sampling mechanisms. We analyse such strategies, but for the Algorithms $1$ of \cite{kovalev2019stochastic}, to improve the theoretical and practical applications of the algorithms. We explore various sampling strategies supporting them with  rigorously constructed experiments.

@@

	\begin{theorem} \label{theorem:1}
		Assume that every $f_i$ is $\mu$-strongly convex and has $H$-Lipschitz Hessians. Then for the random iterates of the Algorithm 1 we have the recursion
		\begin{equation}
			\mathbb E_k[\mathcal W^{k+1}] \leqslant \left( 1 - \frac{\tau}{n} + \frac{\tau}{n} \left( \frac{H}{2\mu} \right)^2 \mathcal W^k \right) \mathcal W^k.
		\end{equation}
		Moreover, if $\forall i\ \|w_i^0 - x^*\| \leqslant \frac{\mu}{H}$ then
		\begin{equation}
			\mathbb E_k[\mathcal W^{k+1}] \leqslant \left( 1 - \frac{3\tau}{4n} \right) \mathcal W^k.
		\end{equation}
	\end{theorem}

@@

	\begin{proof}
		We can generalize the estimation from Lemma \ref{lemma:4} to our case.
		\begin{equation}
			n \cdot \mathbb E_k[\mathcal W^{k+1}] = \sum \limits_{i=1}^n p_i  \| x^{k+1} - x^* \|^2 + \sum \limits_{i=1}^n \left(1 - p_i \right) \|w_i^k - x^*\|^2.
		\end{equation}
		Considering $\tau = \mathbb E_k[|\hat S^k|] = \sum \limits_{i=1}^n p_i$, we have
		\begin{equation}
			n \cdot \mathbb E_k[\mathcal W^{k+1}] = \left( \tau  \| x^{k+1} - x^* \|^2 + n \cdot \mathcal W^k \right) - \sum \limits_{i=1}^n p_i \|w_i^k - x^*\|^2.
		\end{equation}
		So the goal is to maximize $\sum \limits_{i=1}^n p_i \|w_i^k - x^*\|^2$ satisfying the following:
		\begin{equation} \label{polyhedron}
			\begin{cases}
			\sum \limits_{i=1}^n p_i = \tau \\
			0 \leqslant p_i \leqslant 1\ \forall i
			\end{cases}
		\end{equation}
		The solution to this Linear Programming problem is the vertex of the polyhedron defined by the constraints (\ref{polyhedron}). The optimal vertex has the maximum projection of its coordinate vector $(p_1, ..., p_n)^\top$ on the vector $(\|w_i - x^*\|^2, ..., \|w_n - x^*\|^2)$. That is the vertex, which coefficients corresponding to the biggest $\|w_i - x^*\|^2$ are set to 1, corresponding to the smallest $\|w_i - x^*\|^2$ are set to 0, and one coordinate is intermediate if $\tau$ is not an integer.
	\end{proof}

@@

	\subsection{Importance Sampling}
	\begin{figure}[h!]\label{fig:2}
		\centering
		\includegraphics[width=\textwidth]{importance sampling}
		\caption{Comparison of the $\tau$-nice and the Importance Sampling}
	\end{figure}
	We apply both Important Sampling strategies based on Hessians and gradients Lipschitzness parameters. Those are $L_i = \frac{1}{4}\|a_i\|^2$ and $H_i = \frac{1}{10} \|a_i\|^3$ for the logistic regression problem. It is clear that for the considered problem the Importance Sampling used with the Algorithm \ref{SN} does not bring practical improvements.

>>>>

	In practice, various sampling strategies are used for SGD-type algorithms to improve further the performance. One of the most famous sampling mechanisms is so-called Importance Sampling \cite{gower2019sgd, https://doi.org/10.48550/arxiv.1401.2753, 9413313}. The idea is to compute gradients of the functions that have more impact on the problem. Another method of random optimization is Random Reshuffling, which does iterative gradient steps passing reshuffled data \cite{mishchenko2020random}. \cite{richtarik2016parallel} studies many other sampling mechanisms. We analyse such strategies, but for the Algorithms $1$ of \cite{kovalev2019stochastic}, to improve the theoretical and practical applications of the algorithm. We explore various sampling strategies supporting them with  rigorously constructed experiments.

@@

	\begin{theorem} \label{theorem:1}
		Assume that every $f_i$ is $\mu$-strongly convex and has $H$-Lipschitz Hessians. Then for the random iterates of the Algorithm \ref{SN} we have the recursion
		\begin{equation}
			\mathbb E_k[\mathcal W^{k+1}] \leqslant \left( 1 - \frac{\tau}{n} + \frac{\tau}{n} \left( \frac{H}{2\mu} \right)^2 \mathcal W^k \right) \mathcal W^k.
		\end{equation}
		Moreover, if $\forall i\ \|w_i^0 - x^*\| \leqslant \frac{\mu}{H}$ then
		\begin{equation}
			\mathbb E_k[\mathcal W^{k+1}] \leqslant \left( 1 - \frac{3\tau}{4n} \right) \mathcal W^k.
		\end{equation}
	\end{theorem}

@@

	\begin{proof}
		We can generalize the estimation from Lemma \ref{lemma:4} to our case.
		\begin{equation}
			n \cdot \mathbb E_k[\mathcal W^{k+1}] = \sum \limits_{i=1}^n p_i  \| x^{k+1} - x^* \|^2 + \sum \limits_{i=1}^n \left(1 - p_i \right) \|w_i^k - x^*\|^2.
		\end{equation}
		Considering $\tau = \mathbb E_k[|\hat S^k|] = \sum \limits_{i=1}^n p_i$, we have
		\begin{equation}
			n \cdot \mathbb E_k[\mathcal W^{k+1}] = \left( \tau  \| x^{k+1} - x^* \|^2 + n \cdot \mathcal W^k \right) - \sum \limits_{i=1}^n p_i \|w_i^k - x^*\|^2.
		\end{equation}
		So the goal is to maximize $\sum \limits_{i=1}^n p_i \|w_i^k - x^*\|^2$ satisfying the following:
		\begin{equation} \label{polyhedron}
			\begin{cases}
			\sum \limits_{i=1}^n p_i = \tau \\
			0 \leqslant p_i \leqslant 1\ \forall i
			\end{cases}
		\end{equation}
		The solution to this Linear Programming problem is the vertex of the polyhedron defined by the constraints (\ref{polyhedron}). The optimal vertex has the maximum projection of its coordinate vector $(p_1, ..., p_n)^\top$ on the vector $(\|w_i - x^*\|^2, ..., \|w_n - x^*\|^2)$. That is the vertex, whose coefficients corresponding to the biggest $\|w_i - x^*\|^2$ are set to 1, corresponding to the smallest $\|w_i - x^*\|^2$ are set to 0, and one coordinate is intermediate if $\tau$ is not an integer.
	\end{proof}

@@

	\subsection{Importance Sampling}
	\begin{figure}[h!]\label{fig:2}
		\centering
		\includegraphics[width=\textwidth]{importance sampling}
		\caption{Comparison of the $\tau$-nice and the Importance Sampling}
	\end{figure}
	We apply both Important Sampling strategies based on Hessians and gradients Lipschitzness parameters. Those are $H_i = \frac{1}{10} \|a_i\|^3$ and $L_i = \frac{1}{4}\|a_i\|^2$ for the logistic regression problem. It is clear that for the considered problem the Importance Sampling used with the Algorithm \ref{SN} does not bring practical improvements.
